---
title: "Experiment 1b Analysis: Baseline-Sensitive Pairwise Mixed Models"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Overview

This report rewrites the `Experiment 1b` mixed-model analysis around the
baseline-sensitive questions we have been using in the project write-up.

Instead of fitting one omnibus five-level `prime_condition` model and then
recovering the interesting comparisons afterwards, this report fits one direct
pairwise mixed model per planned comparison. That keeps the interpretation
closer to the substantive question.

The pairwise questions are:

1. **Overall prime contrast**
   Does a passive prime yield more passive preference than an active prime?
2. **Baseline bias**
   Do the baseline conditions differ from one another in passive bias?
3. **Active priming relative to baseline**
   Does an active prime increase active preference relative to a given
   baseline?
4. **Passive priming relative to baseline**
   Does a passive prime increase passive preference relative to a given
   baseline?

For every planned comparison, the report fits:

- a binomial GLMM on the binary choice measure
- a linear mixed model on the mean target log-probability difference

To keep the signs interpretable, the dependent variable is direction-specific:

- for **active priming** questions, higher values mean **more active**
  preference
- for **passive priming**, **baseline bias**, and the **active-vs-passive
  prime contrast**, higher values mean **more passive** preference

# Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

required_pkgs <- c(
  "readr", "dplyr", "tidyr", "purrr", "stringr", "forcats",
  "tibble", "glue", "knitr", "lme4", "lmerTest", "emmeans",
  "broom.mixed", "performance"
)

missing_pkgs <- required_pkgs[
  !vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_pkgs) > 0) {
  stop(
    "Install these packages before knitting: ",
    paste(missing_pkgs, collapse = ", ")
  )
}

invisible(lapply(required_pkgs, library, character.only = TRUE))

repo_root <- "/Users/shamiraventurini/PycharmProjects/Geometry-of-Syntax"
exp1_root <- file.path(repo_root, "behavioral_results", "experiment-1")
exp1b_root <- file.path(exp1_root, "experiment-1b")

output_root <- file.path(exp1_root, "r_analysis_outputs")
model_output_root <- file.path(output_root, "mixed_model_outputs_exp1b_pairwise")
dir.create(output_root, recursive = TRUE, showWarnings = FALSE)
dir.create(model_output_root, recursive = TRUE, showWarnings = FALSE)

set.seed(13)

prime_levels <- c("active", "passive", "no_prime_eos", "no_prime_empty", "filler")
baseline_levels <- c("no_prime_eos", "no_prime_empty", "filler")

read_csv_strict <- function(path) {
  if (!file.exists(path)) {
    stop(glue::glue("Missing file: {path}"))
  }
  readr::read_csv(path, show_col_types = FALSE)
}

sanitize_id <- function(x) {
  stringr::str_replace_all(x, "[^A-Za-z0-9_]+", "_")
}
```

# Data Loading

```{r data-loading}
dataset_specs <- tibble::tribble(
  ~dataset_id,                   ~dataset_label,                                 ~model_family, ~variant,            ~condition_key,               ~run_dir,                                                    ~item_dir,
  "gpt2_overlap_core",          "GPT-2 lexical-overlap CORE",                    "gpt2-large",  "lexical_overlap",   "core_core",                  "processing_experiment_1b_gpt2large_v1_lexical-overlap",     "processing_1b_core_core",
  "gpt2_overlap_jabberwocky",   "GPT-2 lexical-overlap Jabberwocky",             "gpt2-large",  "lexical_overlap",   "jabberwocky_jabberwocky",    "processing_experiment_1b_gpt2large_v1_lexical-overlap",     "processing_1b_jabberwocky_jabberwocky",
  "gpt2_strict_core",           "GPT-2 strict-control CORE",                     "gpt2-large",  "strict_control",    "core_core",                  "processing_experiment_1b_gpt2large_v3_strict-control",      "processing_1b_core_core_lexically_controlled",
  "gpt2_strict_jabberwocky",    "GPT-2 strict-control Jabberwocky",              "gpt2-large",  "strict_control",    "jabberwocky_jabberwocky",    "processing_experiment_1b_gpt2large_v3_strict-control",      "processing_1b_jabberwocky_jabberwocky",
  "llama_overlap_core",         "LLaMA lexical-overlap CORE",                    "llama",       "lexical_overlap",   "core_core",                  "processing_experiment_1b_llama_v1_lexical_overlap",         "processing_1b_core_core_lexical_overlap",
  "llama_strict_core",          "LLaMA lexically controlled CORE",               "llama",       "strict_control",    "core_core",                  "processing_experiment_1b_llama_v1_lexically_controlled",    "processing_1b_core_core_lexically_controlled",
  "llama_strict_jabberwocky",   "LLaMA lexically controlled Jabberwocky",        "llama",       "strict_control",    "jabberwocky_jabberwocky",    "processing_experiment_1b_llama_v1_lexically_controlled",    "processing_1b_jabberwocky_jabberwocky"
)

load_dataset_bundle <- function(spec_row) {
  run_root <- file.path(exp1b_root, spec_row$run_dir[[1]])
  item_root <- file.path(run_root, spec_row$item_dir[[1]])

  items <- read_csv_strict(file.path(item_root, "item_scores.csv")) %>%
    mutate(
      item_index = factor(item_index),
      prime_condition = factor(prime_condition, levels = prime_levels),
      passive_choice_indicator = as.numeric(passive_choice_indicator),
      active_choice_indicator = 1 - passive_choice_indicator,
      active_minus_passive_logprob = -passive_minus_active_logprob,
      dataset_id = spec_row$dataset_id[[1]],
      dataset_label = spec_row$dataset_label[[1]],
      model_family = spec_row$model_family[[1]],
      variant = spec_row$variant[[1]],
      condition_key = spec_row$condition_key[[1]]
    )

  list(
    meta = spec_row,
    items = items,
    summary = read_csv_strict(file.path(item_root, "summary.csv")),
    stats = read_csv_strict(file.path(item_root, "stats.csv")),
    priming = read_csv_strict(file.path(run_root, "priming_framed_results.csv"))
  )
}

dataset_bundles <- purrr::map(
  seq_len(nrow(dataset_specs)),
  ~ load_dataset_bundle(dataset_specs[.x, , drop = FALSE])
)
names(dataset_bundles) <- dataset_specs$dataset_id
```

# Descriptive Summaries

```{r descriptive-helpers}
build_overall_summary <- function(bundle) {
  bundle$summary %>%
    select(prime_condition, passive_choice_rate, mean_passive_minus_active_logprob) %>%
    mutate(
      dataset_id = bundle$meta$dataset_id[[1]],
      dataset_label = bundle$meta$dataset_label[[1]],
      active_choice_rate = 1 - passive_choice_rate
    ) %>%
    select(
      dataset_id,
      dataset_label,
      prime_condition,
      passive_choice_rate,
      active_choice_rate,
      mean_passive_minus_active_logprob
    )
}

build_baseline_decomposition <- function(bundle) {
  bundle$priming %>%
    filter(condition == bundle$meta$condition_key[[1]]) %>%
    mutate(
      dataset_id = bundle$meta$dataset_id[[1]],
      dataset_label = bundle$meta$dataset_label[[1]],
      baseline = factor(baseline, levels = baseline_levels)
    ) %>%
    arrange(baseline) %>%
    select(
      dataset_id,
      dataset_label,
      baseline,
      baseline_passive_choice_rate,
      active_choice_priming,
      passive_choice_priming,
      imbalance_choice_passive_minus_active,
      active_logprob_priming,
      passive_logprob_priming,
      imbalance_logprob_passive_minus_active
    )
}

overall_summary_all <- purrr::map_dfr(dataset_bundles, build_overall_summary)
baseline_decomposition_all <- purrr::map_dfr(dataset_bundles, build_baseline_decomposition)

readr::write_csv(overall_summary_all, file.path(output_root, "exp1b_overall_summary_all_runs.csv"))
readr::write_csv(
  baseline_decomposition_all,
  file.path(output_root, "exp1b_baseline_decomposition_all_runs.csv")
)
```

# Planned Pairwise Analysis Specification

```{r planned-comparisons}
analysis_specs <- tibble::tribble(
  ~analysis_id,                              ~analysis_group,          ~reference_condition, ~target_condition, ~direction, ~question,
  "overall_prime_active_vs_passive",         "overall_prime_contrast", "active",             "passive",         "passive",  "Does a passive prime yield more passive preference than an active prime?",
  "baseline_eos_vs_empty",                   "baseline_bias",          "no_prime_eos",       "no_prime_empty",  "passive",  "Is the empty baseline more passive-biased than the EOS baseline?",
  "baseline_eos_vs_filler",                  "baseline_bias",          "no_prime_eos",       "filler",          "passive",  "Is the filler baseline more passive-biased than the EOS baseline?",
  "baseline_empty_vs_filler",                "baseline_bias",          "no_prime_empty",     "filler",          "passive",  "Is the filler baseline more passive-biased than the empty baseline?",
  "active_priming_vs_no_prime_eos",          "active_priming",         "no_prime_eos",       "active",          "active",   "Does an active prime increase active preference relative to the EOS baseline?",
  "active_priming_vs_no_prime_empty",        "active_priming",         "no_prime_empty",     "active",          "active",   "Does an active prime increase active preference relative to the empty baseline?",
  "active_priming_vs_filler",                "active_priming",         "filler",             "active",          "active",   "Does an active prime increase active preference relative to the filler baseline?",
  "passive_priming_vs_no_prime_eos",         "passive_priming",        "no_prime_eos",       "passive",         "passive",  "Does a passive prime increase passive preference relative to the EOS baseline?",
  "passive_priming_vs_no_prime_empty",       "passive_priming",        "no_prime_empty",     "passive",         "passive",  "Does a passive prime increase passive preference relative to the empty baseline?",
  "passive_priming_vs_filler",               "passive_priming",        "filler",             "passive",         "passive",  "Does a passive prime increase passive preference relative to the filler baseline?"
)
```

# Pairwise Model Helpers

```{r model-helpers}
# This helper creates the response variables for the specific question being
# asked. For active-priming models, higher values mean stronger active
# preference. For passive-priming and baseline-bias models, higher values mean
# stronger passive preference.
prepare_pairwise_data <- function(items, reference_condition, target_condition, direction) {
  if (!direction %in% c("active", "passive")) {
    stop(glue::glue("Unsupported direction: {direction}"))
  }

  items %>%
    filter(prime_condition %in% c(reference_condition, target_condition)) %>%
    mutate(
      comparison_condition = factor(
        as.character(prime_condition),
        levels = c(reference_condition, target_condition)
      ),
      response_choice = if (direction == "active") active_choice_indicator else passive_choice_indicator,
      response_logprob = if (direction == "active") active_minus_passive_logprob else passive_minus_active_logprob
    )
}

fit_choice_model <- function(df) {
  tryCatch(
    lme4::glmer(
      response_choice ~ comparison_condition + (1 | item_index),
      data = df,
      family = binomial,
      control = lme4::glmerControl(
        optimizer = "bobyqa",
        optCtrl = list(maxfun = 2e5)
      )
    ),
    error = function(e) e
  )
}

fit_logprob_model <- function(df) {
  tryCatch(
    lmerTest::lmer(
      response_logprob ~ comparison_condition + (1 | item_index),
      data = df
    ),
    error = function(e) e
  )
}

extract_choice_emmeans <- function(model) {
  emm <- as.data.frame(emmeans::emmeans(model, ~ comparison_condition, type = "response"))
  prob_col <- intersect(c("prob", "response"), names(emm))[1]
  ci_low_col <- intersect(c("asymp.LCL", "lower.CL"), names(emm))[1]
  ci_high_col <- intersect(c("asymp.UCL", "upper.CL"), names(emm))[1]
  emm %>%
    mutate(
      emmean = .data[[prob_col]],
      ci_low = .data[[ci_low_col]],
      ci_high = .data[[ci_high_col]]
    ) %>%
    select(comparison_condition, emmean, ci_low, ci_high)
}

extract_logprob_emmeans <- function(model) {
  emm <- as.data.frame(emmeans::emmeans(model, ~ comparison_condition))
  ci_low_col <- intersect(c("lower.CL", "asymp.LCL"), names(emm))[1]
  ci_high_col <- intersect(c("upper.CL", "asymp.UCL"), names(emm))[1]
  emm %>%
    transmute(
      comparison_condition,
      emmean,
      ci_low = .data[[ci_low_col]],
      ci_high = .data[[ci_high_col]]
    )
}

extract_fixed_row <- function(model, target_condition, exponentiate = FALSE) {
  term_name <- paste0("comparison_condition", target_condition)
  broom.mixed::tidy(
    model,
    effects = "fixed",
    conf.int = TRUE,
    conf.method = "Wald",
    exponentiate = exponentiate
  ) %>%
    filter(term == term_name)
}

extract_pair_contrast <- function(model) {
  as.data.frame(
    emmeans::contrast(
      emmeans::emmeans(model, ~ comparison_condition),
      method = "revpairwise"
    )
  )
}

observed_means <- function(df, response_column) {
  df %>%
    group_by(comparison_condition) %>%
    summarise(
      observed_mean = mean(.data[[response_column]], na.rm = TRUE),
      n_items = dplyr::n(),
      .groups = "drop"
    )
}

pull_or_na <- function(df, column) {
  if (!column %in% names(df) || nrow(df) == 0) {
    return(NA_real_)
  }
  value <- df[[column]][[1]]
  if (is.null(value) || length(value) == 0) {
    return(NA_real_)
  }
  as.numeric(value)
}

build_summary_row <- function(
  dataset_id,
  dataset_label,
  analysis_row,
  measure_label,
  observed_df,
  emmeans_df,
  fixed_df,
  model_error = NA_character_
) {
  ref <- analysis_row$reference_condition[[1]]
  tgt <- analysis_row$target_condition[[1]]

  observed_ref <- observed_df %>% filter(comparison_condition == ref)
  observed_tgt <- observed_df %>% filter(comparison_condition == tgt)
  emm_ref <- emmeans_df %>% filter(comparison_condition == ref)
  emm_tgt <- emmeans_df %>% filter(comparison_condition == tgt)

  tibble::tibble(
    dataset_id = dataset_id,
    dataset_label = dataset_label,
    analysis_id = analysis_row$analysis_id[[1]],
    analysis_group = analysis_row$analysis_group[[1]],
    measure = measure_label,
    direction = analysis_row$direction[[1]],
    question = analysis_row$question[[1]],
    reference_condition = ref,
    target_condition = tgt,
    observed_reference = observed_ref$observed_mean[[1]],
    observed_target = observed_tgt$observed_mean[[1]],
    observed_target_minus_reference = observed_tgt$observed_mean[[1]] - observed_ref$observed_mean[[1]],
    emmean_reference = emm_ref$emmean[[1]],
    emmean_target = emm_tgt$emmean[[1]],
    emmean_reference_ci_low = emm_ref$ci_low[[1]],
    emmean_reference_ci_high = emm_ref$ci_high[[1]],
    emmean_target_ci_low = emm_tgt$ci_low[[1]],
    emmean_target_ci_high = emm_tgt$ci_high[[1]],
    fixed_effect_estimate = pull_or_na(fixed_df, "estimate"),
    fixed_effect_ci_low = pull_or_na(fixed_df, "conf.low"),
    fixed_effect_ci_high = pull_or_na(fixed_df, "conf.high"),
    fixed_effect_statistic = pull_or_na(fixed_df, "statistic"),
    fixed_effect_p = pull_or_na(fixed_df, "p.value"),
    model_error = model_error
  )
}

run_pairwise_analysis <- function(bundle, analysis_row) {
  dataset_id <- bundle$meta$dataset_id[[1]]
  dataset_label <- bundle$meta$dataset_label[[1]]
  analysis_id <- analysis_row$analysis_id[[1]]
  dataset_output_dir <- file.path(model_output_root, dataset_id)
  dir.create(dataset_output_dir, recursive = TRUE, showWarnings = FALSE)

  pair_df <- prepare_pairwise_data(
    items = bundle$items,
    reference_condition = analysis_row$reference_condition[[1]],
    target_condition = analysis_row$target_condition[[1]],
    direction = analysis_row$direction[[1]]
  )

  file_stub <- sanitize_id(analysis_id)

  choice_model <- fit_choice_model(pair_df)
  if (inherits(choice_model, "error")) {
    choice_error <- conditionMessage(choice_model)
    choice_fixed <- tibble::tibble(term = NA_character_, estimate = NA_real_, conf.low = NA_real_, conf.high = NA_real_, statistic = NA_real_, p.value = NA_real_)
    choice_emm <- tibble::tibble(comparison_condition = levels(pair_df$comparison_condition), emmean = NA_real_, ci_low = NA_real_, ci_high = NA_real_)
    choice_contrast <- tibble::tibble(contrast = NA_character_, estimate = NA_real_, SE = NA_real_, z.ratio = NA_real_, p.value = NA_real_)
  } else {
    choice_error <- NA_character_
    choice_fixed <- extract_fixed_row(choice_model, analysis_row$target_condition[[1]], exponentiate = FALSE)
    choice_emm <- extract_choice_emmeans(choice_model)
    choice_contrast <- extract_pair_contrast(choice_model)
  }

  logprob_model <- fit_logprob_model(pair_df)
  if (inherits(logprob_model, "error")) {
    logprob_error <- conditionMessage(logprob_model)
    logprob_fixed <- tibble::tibble(term = NA_character_, estimate = NA_real_, conf.low = NA_real_, conf.high = NA_real_, statistic = NA_real_, p.value = NA_real_)
    logprob_emm <- tibble::tibble(comparison_condition = levels(pair_df$comparison_condition), emmean = NA_real_, ci_low = NA_real_, ci_high = NA_real_)
    logprob_contrast <- tibble::tibble(contrast = NA_character_, estimate = NA_real_, SE = NA_real_, df = NA_real_, t.ratio = NA_real_, p.value = NA_real_)
  } else {
    logprob_error <- NA_character_
    logprob_fixed <- extract_fixed_row(logprob_model, analysis_row$target_condition[[1]], exponentiate = FALSE)
    logprob_emm <- extract_logprob_emmeans(logprob_model)
    logprob_contrast <- extract_pair_contrast(logprob_model)
  }

  choice_observed <- observed_means(pair_df, "response_choice")
  logprob_observed <- observed_means(pair_df, "response_logprob")

  readr::write_csv(pair_df, file.path(dataset_output_dir, paste0(file_stub, "_analysis_data.csv")))
  readr::write_csv(choice_observed, file.path(dataset_output_dir, paste0(file_stub, "_choice_observed_means.csv")))
  readr::write_csv(choice_emm, file.path(dataset_output_dir, paste0(file_stub, "_choice_emmeans.csv")))
  readr::write_csv(choice_contrast, file.path(dataset_output_dir, paste0(file_stub, "_choice_posthoc_contrast.csv")))
  readr::write_csv(choice_fixed, file.path(dataset_output_dir, paste0(file_stub, "_choice_fixed_effect.csv")))
  readr::write_csv(logprob_observed, file.path(dataset_output_dir, paste0(file_stub, "_logprob_observed_means.csv")))
  readr::write_csv(logprob_emm, file.path(dataset_output_dir, paste0(file_stub, "_logprob_emmeans.csv")))
  readr::write_csv(logprob_contrast, file.path(dataset_output_dir, paste0(file_stub, "_logprob_posthoc_contrast.csv")))
  readr::write_csv(logprob_fixed, file.path(dataset_output_dir, paste0(file_stub, "_logprob_fixed_effect.csv")))

  list(
    choice_summary = build_summary_row(
      dataset_id = dataset_id,
      dataset_label = dataset_label,
      analysis_row = analysis_row,
      measure_label = "choice",
      observed_df = choice_observed,
      emmeans_df = choice_emm,
      fixed_df = choice_fixed,
      model_error = choice_error
    ),
    logprob_summary = build_summary_row(
      dataset_id = dataset_id,
      dataset_label = dataset_label,
      analysis_row = analysis_row,
      measure_label = "logprob",
      observed_df = logprob_observed,
      emmeans_df = logprob_emm,
      fixed_df = logprob_fixed,
      model_error = logprob_error
    )
  )
}

run_dataset_suite <- function(bundle) {
  analysis_results <- purrr::map(
    seq_len(nrow(analysis_specs)),
    ~ run_pairwise_analysis(bundle, analysis_specs[.x, , drop = FALSE])
  )

  choice_summary <- purrr::map_dfr(analysis_results, "choice_summary")
  logprob_summary <- purrr::map_dfr(analysis_results, "logprob_summary")
  overall_summary <- build_overall_summary(bundle)
  baseline_summary <- build_baseline_decomposition(bundle)

  dataset_output_dir <- file.path(model_output_root, bundle$meta$dataset_id[[1]])
  dir.create(dataset_output_dir, recursive = TRUE, showWarnings = FALSE)

  readr::write_csv(overall_summary, file.path(dataset_output_dir, "stored_overall_summary.csv"))
  readr::write_csv(baseline_summary, file.path(dataset_output_dir, "stored_baseline_decomposition.csv"))
  readr::write_csv(choice_summary, file.path(dataset_output_dir, "pairwise_choice_model_summary.csv"))
  readr::write_csv(logprob_summary, file.path(dataset_output_dir, "pairwise_logprob_model_summary.csv"))

  list(
    overall_summary = overall_summary,
    baseline_summary = baseline_summary,
    choice_summary = choice_summary,
    logprob_summary = logprob_summary
  )
}

dataset_results <- purrr::map(dataset_bundles, run_dataset_suite)
```

# Pairwise Mixed-Model Reports

```{r dataset-report, results='asis'}
for (dataset_id in names(dataset_results)) {
  bundle <- dataset_bundles[[dataset_id]]
  result <- dataset_results[[dataset_id]]

  cat("## ", bundle$meta$dataset_label[[1]], "\n\n", sep = "")
  cat("Run directory: `", bundle$meta$run_dir[[1]], "`  \n", sep = "")
  cat("Condition key: `", bundle$meta$condition_key[[1]], "`  \n", sep = "")
  cat("Model family: `", bundle$meta$model_family[[1]], "`  \n", sep = "")
  cat("Variant: `", bundle$meta$variant[[1]], "`\n\n", sep = "")

  cat("### Stored Overall Summary\n\n")
  print(knitr::kable(result$overall_summary, digits = 4))
  cat("\n\n")

  cat("### Stored Baseline Decomposition\n\n")
  print(knitr::kable(result$baseline_summary, digits = 4))
  cat("\n\n")

  cat("### Planned Pairwise Mixed Models: Choice Metric\n\n")
  cat("Each row below is one direct GLMM. The sign and the observed means are\n")
  cat("already oriented so that larger values mean more of the structure named in\n")
  cat("the `direction` column.\n\n")
  print(
    knitr::kable(
      result$choice_summary %>%
        select(
          analysis_group, analysis_id, direction, reference_condition,
          target_condition, observed_reference, observed_target,
          observed_target_minus_reference, fixed_effect_estimate,
          fixed_effect_p, model_error
        ),
      digits = 4
    )
  )
  cat("\n\n")

  cat("### Planned Pairwise Mixed Models: Log-Probability Metric\n\n")
  cat("Each row below is one direct LMM on the directional log-probability score.\n\n")
  print(
    knitr::kable(
      result$logprob_summary %>%
        select(
          analysis_group, analysis_id, direction, reference_condition,
          target_condition, observed_reference, observed_target,
          observed_target_minus_reference, fixed_effect_estimate,
          fixed_effect_p, model_error
        ),
      digits = 4
    )
  )
  cat("\n\n")
}
```

# Combined Summary Tables

```{r combined-summary-exports}
pairwise_choice_all <- purrr::map_dfr(dataset_results, "choice_summary")
pairwise_logprob_all <- purrr::map_dfr(dataset_results, "logprob_summary")

readr::write_csv(
  pairwise_choice_all,
  file.path(output_root, "exp1b_pairwise_choice_models_all_runs.csv")
)
readr::write_csv(
  pairwise_logprob_all,
  file.path(output_root, "exp1b_pairwise_logprob_models_all_runs.csv")
)

knitr::kable(
  pairwise_choice_all %>%
    select(
      dataset_label, analysis_id, direction,
      reference_condition, target_condition,
      observed_target_minus_reference,
      fixed_effect_estimate, fixed_effect_p
    ),
  digits = 4
)
```

# Files Written by This Report

```{r generated-files}
tibble::tibble(
  file = list.files(model_output_root, recursive = TRUE)
) %>%
  knitr::kable()
```
