---
title: "Structural Priming Suite: Mixed-Effects Analysis Plan"
author: "Geometry of Syntax"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
params:
  repo_root: "."
  input_dir: "behavioral_results/analysis_ready"
  output_dir: "behavioral_results/r_mixed_models"
  item_level_path: "item_level_master.csv"
  baseline_shifts_path: "baseline_centered_logodds_shifts.csv"
  roi_long_path: "roi_pe_item_level.csv"
  roi_percentage_path: "roi_pe_percentage_summary.csv"
  ife_path: "ife_prime_surprisal_item_level.csv"
  exp2_generation_path: "exp2_generation_item_level.csv"
  exp4_comprehension_path: "exp4_comprehension_item_level.csv"
  model_size_reference: "small"
  lexicality_reference: "core"
  baseline_reference: "no_prime"
  run_sensitivity_models: true
  save_model_objects: false
---

# Purpose

This notebook is the analysis scaffold for the active/passive structural priming
suite. It is organized by **research question**, not by experiment. The same
logic is then applied to whichever experiments and models are present in the
analysis-ready CSVs.

The notebook assumes that Python-side scripts have already exported tidy,
analysis-ready files. If a file is missing, the relevant section is skipped with
a clear message rather than failing the whole notebook.

Primary analysis principle:

- report baseline active/passive preference separately from prime-induced shift;
- use `model_size` as the primary model comparison variable;
- treat `model_family` as descriptive or robustness-only;
- include item/order/token controls wherever available;
- use item-level random intercepts where repeated target items are observed.

# Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

required_pkgs <- c(
  "readr", "dplyr", "tidyr", "purrr", "stringr", "forcats",
  "ggplot2", "tibble", "glue", "knitr",
  "lme4", "emmeans", "broom.mixed", "performance"
)

missing_pkgs <- required_pkgs[
  !vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_pkgs) > 0) {
  stop("Install these packages before knitting: ", paste(missing_pkgs, collapse = ", "))
}

invisible(lapply(required_pkgs, library, character.only = TRUE))

repo_root <- normalizePath(params$repo_root, mustWork = FALSE)
input_dir <- file.path(repo_root, params$input_dir)
output_dir <- file.path(repo_root, params$output_dir)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

set.seed(13)

theme_set(ggplot2::theme_minimal(base_size = 12))
```

# Helpers

```{r helpers}
path_from_param <- function(param_value) {
  path <- as.character(param_value)
  if (grepl("^(/|[A-Za-z]:[/\\\\])", path)) {
    return(path)
  }
  file.path(input_dir, path)
}
```

```{r helper-functions}
read_optional_csv <- function(path, label) {
  if (!file.exists(path)) {
    message(glue::glue("Skipping {label}; missing file: {path}"))
    return(NULL)
  }
  readr::read_csv(path, show_col_types = FALSE)
}

has_cols <- function(df, cols) {
  !is.null(df) && all(cols %in% names(df))
}

require_cols <- function(df, cols, label) {
  missing <- setdiff(cols, names(df))
  if (length(missing) > 0) {
    stop(glue::glue("{label} is missing required columns: {paste(missing, collapse = ', ')}"))
  }
  invisible(df)
}

first_present <- function(df, candidates) {
  found <- candidates[candidates %in% names(df)]
  if (length(found) == 0) {
    return(NULL)
  }
  found[[1]]
}

z_if_present <- function(df, col) {
  if (!col %in% names(df)) {
    return(df)
  }
  out <- df
  new_col <- paste0(col, "_z")
  out[[new_col]] <- as.numeric(scale(out[[col]]))
  out
}

coalesce_missing_col <- function(df, col, value) {
  if (!col %in% names(df)) {
    df[[col]] <- value
  }
  df
}

canonicalize_common <- function(df) {
  if (is.null(df)) {
    return(NULL)
  }

  out <- df %>%
    mutate(
      prime_condition = if ("prime_condition" %in% names(.)) {
        stringr::str_replace_all(as.character(prime_condition), c(
          "active_prime" = "active",
          "passive_prime" = "passive",
          "filler_prime" = "filler",
          "no_prime_eos" = "no_prime",
          "no_prime_empty" = "no_prime",
          "no_demo" = "no_prime"
        ))
      } else {
        NA_character_
      }
    )

  out <- coalesce_missing_col(out, "role_order", "not_applicable")
  out <- coalesce_missing_col(out, "task_type", "unspecified")
  out <- coalesce_missing_col(out, "model_size", "unspecified")
  out <- coalesce_missing_col(out, "model_family", "unspecified")
  out <- coalesce_missing_col(out, "lexicality_condition", "unspecified")
  out <- coalesce_missing_col(out, "item_id", NA_character_)
  out <- coalesce_missing_col(out, "item_index", NA_integer_)

  out %>%
    mutate(
      prime_condition = factor(
        prime_condition,
        levels = c("no_prime", "filler", "active", "passive")
      ),
      lexicality_condition = forcats::fct_relevel(
        factor(lexicality_condition),
        params$lexicality_reference
      ),
      model_size = forcats::fct_relevel(
        factor(model_size),
        params$model_size_reference
      ),
      model_family = factor(model_family),
      task_type = factor(task_type),
      role_order = factor(role_order),
      item_factor = factor(ifelse(is.na(item_id), item_index, item_id))
    ) %>%
    z_if_present("item_order") %>%
    z_if_present("item_index") %>%
    z_if_present("token_count_diff") %>%
    z_if_present("passive_minus_active_token_count") %>%
    z_if_present("prime_token_count_diff") %>%
    z_if_present("target_token_count_diff")
}

available_controls <- function(df, candidates) {
  candidates[candidates %in% names(df)]
}

random_item_term <- function(df) {
  if ("item_factor" %in% names(df) && dplyr::n_distinct(df$item_factor, na.rm = TRUE) > 1) {
    "(1 | item_factor)"
  } else {
    NULL
  }
}

build_formula <- function(response, fixed_terms, random_terms = NULL) {
  fixed_terms <- fixed_terms[!is.na(fixed_terms) & nzchar(fixed_terms)]
  random_terms <- random_terms[!is.na(random_terms) & nzchar(random_terms)]
  rhs <- c(fixed_terms, random_terms)
  stats::as.formula(paste(response, "~", paste(rhs, collapse = " + ")))
}

fit_lmer_safe <- function(formula, data, label) {
  message(glue::glue("Fitting LMM: {label}"))
  message(format(formula))
  tryCatch(
    lme4::lmer(
      formula,
      data = data,
      REML = FALSE,
      control = lme4::lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
    ),
    error = function(e) {
      warning(glue::glue("Model failed [{label}]: {conditionMessage(e)}"))
      NULL
    }
  )
}

fit_glmer_safe <- function(formula, data, label) {
  message(glue::glue("Fitting GLMM: {label}"))
  message(format(formula))
  tryCatch(
    lme4::glmer(
      formula,
      data = data,
      family = binomial,
      control = lme4::glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
    ),
    error = function(e) {
      warning(glue::glue("Model failed [{label}]: {conditionMessage(e)}"))
      NULL
    }
  )
}

tidy_model <- function(model, model_id) {
  if (is.null(model)) {
    return(tibble::tibble())
  }
  broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
    mutate(model_id = model_id, .before = 1)
}

save_model_summary <- function(model, model_id) {
  if (is.null(model)) {
    return(invisible(NULL))
  }
  txt_path <- file.path(output_dir, paste0(model_id, "_summary.txt"))
  capture.output(summary(model), file = txt_path)

  if (isTRUE(params$save_model_objects)) {
    saveRDS(model, file.path(output_dir, paste0(model_id, ".rds")))
  }
  invisible(NULL)
}

write_model_table <- function(rows, filename) {
  table <- dplyr::bind_rows(rows)
  if (nrow(table) > 0) {
    readr::write_csv(table, file.path(output_dir, filename))
  }
  table
}
```

# Load Data

Expected analysis-ready files:

- `item_level_master.csv`: item-level log-probability and/or choice measures.
- `baseline_centered_logodds_shifts.csv`: one row per item/model/baseline-centered shift.
- `roi_pe_item_level.csv`: long ROI-level PE table.
- `roi_pe_percentage_summary.csv`: region percentage summaries.
- `ife_prime_surprisal_item_level.csv`: IFE-style surprisal and target shifts.
- `exp2_generation_item_level.csv`: generation labels for Experiment 2, if not included in master.
- `exp4_comprehension_item_level.csv`: comprehension/role-recovery accuracy for Experiment 4, if not included in master.

```{r load-data}
item_level <- read_optional_csv(path_from_param(params$item_level_path), "item-level master") %>%
  canonicalize_common()

baseline_shifts <- read_optional_csv(path_from_param(params$baseline_shifts_path), "baseline-centered shifts") %>%
  canonicalize_common()

roi_long <- read_optional_csv(path_from_param(params$roi_long_path), "ROI item-level PE") %>%
  canonicalize_common()

roi_percentage <- read_optional_csv(path_from_param(params$roi_percentage_path), "ROI percentage summary")

ife_item <- read_optional_csv(path_from_param(params$ife_path), "IFE item-level") %>%
  canonicalize_common()

exp2_generation <- read_optional_csv(path_from_param(params$exp2_generation_path), "Experiment 2 generation item-level") %>%
  canonicalize_common()

exp4_comprehension <- read_optional_csv(path_from_param(params$exp4_comprehension_path), "Experiment 4 comprehension item-level") %>%
  canonicalize_common()
```

# Design Checks

```{r design-checks}
describe_table <- function(df, label) {
  if (is.null(df)) {
    return(invisible(NULL))
  }
  cat("\n\n##", label, "\n")
  print(dim(df))
  print(names(df))
}

describe_table(item_level, "item_level")
describe_table(baseline_shifts, "baseline_shifts")
describe_table(roi_long, "roi_long")
describe_table(ife_item, "ife_item")
describe_table(exp2_generation, "exp2_generation")
describe_table(exp4_comprehension, "exp4_comprehension")
```

```{r balance-checks}
if (has_cols(item_level, c("experiment", "task_type", "lexicality_condition", "model_size", "prime_condition"))) {
  item_level %>%
    count(experiment, task_type, lexicality_condition, model_size, prime_condition) %>%
    arrange(experiment, task_type, lexicality_condition, model_size, prime_condition) %>%
    knitr::kable()
} else {
  message("Skipping balance table; item_level is missing or lacks required grouping columns.")
}
```

# Descriptive Plots

```{r descriptive-plots, fig.width=9, fig.height=5}
if (has_cols(item_level, c("logodds_passive", "prime_condition", "lexicality_condition", "model_size"))) {
  ggplot(item_level, aes(prime_condition, logodds_passive, fill = lexicality_condition)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    stat_summary(fun = mean, geom = "col", position = position_dodge(width = 0.8)) +
    stat_summary(fun.data = mean_cl_boot, geom = "errorbar", position = position_dodge(width = 0.8), width = 0.2) +
    facet_wrap(~ model_size) +
    labs(
      title = "Passive log-odds by prime condition",
      y = "logP(passive target) - logP(active target)",
      x = "Prime condition"
    )
}
```

# RQ1: Baseline Active/Passive Bias

Question: before priming, do models prefer passive or active targets, and does
this vary by lexicality, task, and model size?

Primary DV:

```text
logodds_passive = logP(passive target) - logP(active target)
```

Primary subset: `prime_condition == no_prime`. Filler baseline is a robustness
check, not the main baseline.

```{r rq1-baseline-bias}
rq1_models <- list()

if (has_cols(item_level, c("logodds_passive", "prime_condition", "lexicality_condition", "model_size"))) {
  baseline_df <- item_level %>%
    filter(prime_condition == params$baseline_reference) %>%
    filter(!is.na(logodds_passive))

  controls <- available_controls(
    baseline_df,
    c("task_type", "token_count_diff_z", "passive_minus_active_token_count_z", "item_order_z", "item_index_z")
  )

  fixed <- c("lexicality_condition * model_size", controls)
  formula <- build_formula("logodds_passive", fixed, random_item_term(baseline_df))

  rq1_models$baseline_bias <- fit_lmer_safe(formula, baseline_df, "rq1_baseline_bias")
  save_model_summary(rq1_models$baseline_bias, "rq1_baseline_bias")
}

rq1_table <- write_model_table(
  purrr::imap(rq1_models, ~ tidy_model(.x, .y)),
  "rq1_baseline_bias_fixed_effects.csv"
)
rq1_table %>% knitr::kable()
```

```{r rq1-filler-sensitivity}
if (isTRUE(params$run_sensitivity_models) &&
    has_cols(item_level, c("logodds_passive", "prime_condition", "lexicality_condition", "model_size"))) {
  filler_df <- item_level %>%
    filter(prime_condition == "filler") %>%
    filter(!is.na(logodds_passive))

  if (nrow(filler_df) > 0) {
    controls <- available_controls(
      filler_df,
      c("task_type", "token_count_diff_z", "passive_minus_active_token_count_z", "item_order_z", "item_index_z")
    )
    formula <- build_formula(
      "logodds_passive",
      c("lexicality_condition * model_size", controls),
      random_item_term(filler_df)
    )
    rq1_filler <- fit_lmer_safe(formula, filler_df, "rq1_filler_baseline_sensitivity")
    save_model_summary(rq1_filler, "rq1_filler_baseline_sensitivity")
    tidy_model(rq1_filler, "rq1_filler_baseline_sensitivity") %>% knitr::kable()
  }
}
```

# RQ2: Baseline-Centered Priming Shift

Question: does an active or passive prime move the model away from its own
no-prime baseline?

Recommended table: `baseline_centered_logodds_shifts.csv`.

Expected columns:

```text
shift_from_no_prime
prime_type              # active / passive
lexicality_condition
task_type
model_size
item_id
token_count_diff
role_order
item_order
```

```{r rq2-baseline-centered-shift}
rq2_models <- list()

if (has_cols(baseline_shifts, c("shift_from_no_prime", "prime_type", "lexicality_condition", "model_size"))) {
  shift_df <- baseline_shifts %>%
    mutate(
      prime_type = factor(prime_type, levels = c("active", "passive"))
    ) %>%
    filter(!is.na(shift_from_no_prime))

  controls <- available_controls(
    shift_df,
    c("task_type", "role_order", "token_count_diff_z", "target_token_count_diff_z", "prime_token_count_diff_z", "item_order_z", "item_index_z")
  )

  formula <- build_formula(
    "shift_from_no_prime",
    c("prime_type * lexicality_condition * model_size", controls),
    random_item_term(shift_df)
  )

  rq2_models$shift_no_prime <- fit_lmer_safe(formula, shift_df, "rq2_shift_from_no_prime")
  save_model_summary(rq2_models$shift_no_prime, "rq2_shift_from_no_prime")
}

rq2_table <- write_model_table(
  purrr::imap(rq2_models, ~ tidy_model(.x, .y)),
  "rq2_baseline_centered_shift_fixed_effects.csv"
)
rq2_table %>% knitr::kable()
```

```{r rq2-estimated-marginal-means}
if (!is.null(rq2_models$shift_no_prime)) {
  emmeans::emmeans(
    rq2_models$shift_no_prime,
    specs = ~ prime_type | lexicality_condition * model_size
  ) %>%
    as.data.frame() %>%
    readr::write_csv(file.path(output_dir, "rq2_shift_emmeans.csv"))

  emmeans::contrast(
    emmeans::emmeans(rq2_models$shift_no_prime, specs = ~ prime_type | lexicality_condition * model_size),
    method = "pairwise"
  ) %>%
    as.data.frame() %>%
    readr::write_csv(file.path(output_dir, "rq2_active_vs_passive_shift_contrasts.csv"))
}
```

# RQ3: Does Priming Survive in Jabberwocky?

Question: does the prime-induced shift persist when lexical semantics are
removed/replaced by the Jabberwocky condition?

This is technically part of RQ2, but it deserves a focused contrast because it
is central to interpretation.

```{r rq3-jabberwocky}
rq3_models <- list()

if (has_cols(baseline_shifts, c("shift_from_no_prime", "prime_type", "lexicality_condition", "model_size"))) {
  jab_df <- baseline_shifts %>%
    mutate(prime_type = factor(prime_type, levels = c("active", "passive"))) %>%
    filter(lexicality_condition %in% c("core", "jabberwocky", "nonce", "real")) %>%
    filter(!is.na(shift_from_no_prime))

  controls <- available_controls(
    jab_df,
    c("task_type", "role_order", "token_count_diff_z", "target_token_count_diff_z", "prime_token_count_diff_z", "item_order_z", "item_index_z")
  )

  formula <- build_formula(
    "shift_from_no_prime",
    c("prime_type * lexicality_condition * model_size", controls),
    random_item_term(jab_df)
  )

  rq3_models$lexicality_shift <- fit_lmer_safe(formula, jab_df, "rq3_lexicality_shift")
  save_model_summary(rq3_models$lexicality_shift, "rq3_lexicality_shift")
}

rq3_table <- write_model_table(
  purrr::imap(rq3_models, ~ tidy_model(.x, .y)),
  "rq3_jabberwocky_fixed_effects.csv"
)
rq3_table %>% knitr::kable()
```

# RQ4: Does the Effect Generalize Across Tasks?

Question: is the same active/passive priming pattern visible in fixed-target
processing, free generation, event-to-syntax mapping, and comprehension?

This model only makes sense after the dependent variables have been put onto a
comparable scale, e.g. baseline-centered shift or standardized effect.

```{r rq4-task-generalization}
rq4_models <- list()

task_effect_col <- NULL
if (!is.null(baseline_shifts)) {
  task_effect_col <- first_present(baseline_shifts, c("effect_z", "shift_z", "shift_from_no_prime"))
}

if (!is.null(task_effect_col) &&
    has_cols(baseline_shifts, c(task_effect_col, "prime_type", "task_type", "lexicality_condition", "model_size"))) {
  task_df <- baseline_shifts %>%
    mutate(
      prime_type = factor(prime_type, levels = c("active", "passive")),
      effect_for_model = .data[[task_effect_col]]
    ) %>%
    filter(!is.na(effect_for_model))

  controls <- available_controls(
    task_df,
    c("role_order", "token_count_diff_z", "target_token_count_diff_z", "prime_token_count_diff_z", "item_order_z", "item_index_z")
  )

  formula <- build_formula(
    "effect_for_model",
    c("prime_type * lexicality_condition * task_type * model_size", controls),
    random_item_term(task_df)
  )

  rq4_models$task_generalization <- fit_lmer_safe(formula, task_df, "rq4_task_generalization")
  save_model_summary(rq4_models$task_generalization, "rq4_task_generalization")
}

rq4_table <- write_model_table(
  purrr::imap(rq4_models, ~ tidy_model(.x, .y)),
  "rq4_task_generalization_fixed_effects.csv"
)
rq4_table %>% knitr::kable()
```

# Experiment 2: Generation Readout

Question: does the prime change the structure the model chooses to generate?

Primary DV: binary passive generation.

```{r exp2-generation-glmm}
exp2_models <- list()

if (has_cols(exp2_generation, c("generation_voice", "prime_condition", "lexicality_condition", "model_size"))) {
  exp2_df <- exp2_generation %>%
    filter(generation_voice %in% c("active", "passive")) %>%
    mutate(
      is_passive = as.integer(generation_voice == "passive")
    )

  controls <- available_controls(
    exp2_df,
    c("role_order", "target_verb_cue", "item_order_z", "item_index_z")
  )

  formula <- build_formula(
    "is_passive",
    c("prime_condition * lexicality_condition * model_size", controls),
    random_item_term(exp2_df)
  )

  exp2_models$generation_passive <- fit_glmer_safe(formula, exp2_df, "exp2_generation_passive")
  save_model_summary(exp2_models$generation_passive, "exp2_generation_passive")
}

exp2_table <- write_model_table(
  purrr::imap(exp2_models, ~ tidy_model(.x, .y)),
  "exp2_generation_fixed_effects.csv"
)
exp2_table %>% knitr::kable()
```

# Experiment 4: Comprehension / Role Recovery

Question: does priming or target voice affect role-recovery accuracy?

Primary DV: correct answer to doer / acted-on question.

```{r exp4-comprehension-glmm}
exp4_models <- list()

if (has_cols(exp4_comprehension, c("is_correct", "prime_condition", "target_voice", "question_type", "lexicality_condition", "model_size"))) {
  exp4_df <- exp4_comprehension %>%
    mutate(
      is_correct = as.integer(is_correct),
      target_voice = factor(target_voice, levels = c("active", "passive")),
      question_type = factor(question_type, levels = c("doer", "acted_on"))
    )

  controls <- available_controls(
    exp4_df,
    c("complexity_condition", "item_order_z", "item_index_z")
  )

  formula <- build_formula(
    "is_correct",
    c("prime_condition * target_voice * question_type * lexicality_condition * model_size", controls),
    random_item_term(exp4_df)
  )

  exp4_models$role_recovery_accuracy <- fit_glmer_safe(formula, exp4_df, "exp4_role_recovery_accuracy")
  save_model_summary(exp4_models$role_recovery_accuracy, "exp4_role_recovery_accuracy")
}

exp4_table <- write_model_table(
  purrr::imap(exp4_models, ~ tidy_model(.x, .y)),
  "exp4_comprehension_fixed_effects.csv"
)
exp4_table %>% knitr::kable()
```

# ROI Analysis

Question: where in the sentence does the priming effect occur?

Primary DV: ROI-level PE or log-probability difference.

```{r roi-model}
roi_models <- list()

roi_dv <- NULL
if (!is.null(roi_long)) {
  roi_dv <- first_present(roi_long, c("roi_pe", "pe", "roi_logprob_delta", "logprob_delta"))
}

if (!is.null(roi_dv) &&
    has_cols(roi_long, c(roi_dv, "roi", "prime_condition", "target_voice", "lexicality_condition", "model_size"))) {
  roi_df <- roi_long %>%
    mutate(
      roi = factor(roi),
      target_voice = factor(target_voice, levels = c("active", "passive")),
      roi_response = .data[[roi_dv]]
    ) %>%
    filter(!is.na(roi_response))

  controls <- available_controls(
    roi_df,
    c("roi_token_count", "item_order_z", "item_index_z")
  )

  formula <- build_formula(
    "roi_response",
    c("roi * prime_condition * lexicality_condition * model_size", "target_voice", controls),
    random_item_term(roi_df)
  )

  roi_models$roi_pe <- fit_lmer_safe(formula, roi_df, "roi_pe")
  save_model_summary(roi_models$roi_pe, "roi_pe")
}

roi_table <- write_model_table(
  purrr::imap(roi_models, ~ tidy_model(.x, .y)),
  "roi_pe_fixed_effects.csv"
)
roi_table %>% knitr::kable()
```

```{r roi-percentages}
if (!is.null(roi_percentage)) {
  readr::write_csv(roi_percentage, file.path(output_dir, "roi_pe_percentage_summary_copy.csv"))
  roi_percentage %>% head(30) %>% knitr::kable()
}
```

# IFE-Style Prime Surprisal Analysis

Question: do less expected primes induce larger target shifts?

Primary DV: baseline-centered target shift.

Primary predictor:

```text
prime_surprisal = -log P(prime structure at baseline)
```

```{r ife-model}
ife_models <- list()

ife_dv <- NULL
if (!is.null(ife_item)) {
  ife_dv <- first_present(ife_item, c("target_shift", "shift_from_no_prime", "passive_target_shift", "active_target_shift"))
}

if (!is.null(ife_dv) &&
    has_cols(ife_item, c(ife_dv, "prime_surprisal", "prime_type", "lexicality_condition", "model_size"))) {
  ife_df <- ife_item %>%
    mutate(
      prime_type = factor(prime_type, levels = c("active", "passive")),
      ife_response = .data[[ife_dv]]
    ) %>%
    filter(!is.na(ife_response), !is.na(prime_surprisal))

  controls <- available_controls(
    ife_df,
    c("role_order", "token_count_diff_z", "target_token_count_diff_z", "prime_token_count_diff_z", "item_order_z", "item_index_z")
  )

  formula <- build_formula(
    "ife_response",
    c("prime_surprisal * prime_type * lexicality_condition * model_size", controls),
    random_item_term(ife_df)
  )

  ife_models$prime_surprisal <- fit_lmer_safe(formula, ife_df, "ife_prime_surprisal")
  save_model_summary(ife_models$prime_surprisal, "ife_prime_surprisal")
}

ife_table <- write_model_table(
  purrr::imap(ife_models, ~ tidy_model(.x, .y)),
  "ife_prime_surprisal_fixed_effects.csv"
)
ife_table %>% knitr::kable()
```

# Sensitivity Models

These should not replace the primary models. They answer: does the conclusion
survive reasonable alternative specifications?

Planned sensitivity dimensions:

- total-logprob vs mean-token logprob;
- no-prime baseline vs filler baseline;
- with vs without token-count covariates;
- with vs without `role_order`;
- with vs without centered item order;
- per-model descriptive replications.

```{r sensitivity-notes}
sensitivity_plan <- tibble::tribble(
  ~sensitivity, ~purpose,
  "mean-token logprob", "Reduce length sensitivity in cross-corpus comparisons.",
  "total-logprob", "Preserve whole-candidate preference as a Sinclair-style score.",
  "filler baseline", "Check whether results depend on no-prime baseline only.",
  "drop token-count covariates", "Check whether tokenization controls drive the effect.",
  "drop role_order", "Check whether deterministic prompt-role ordering drives the effect.",
  "per-model fits", "Avoid over-aggregating across model families/tokenizers."
)

sensitivity_plan %>% knitr::kable()
```

# Model Diagnostics

```{r diagnostics}
all_models <- c(rq1_models, rq2_models, rq3_models, rq4_models, exp2_models, exp4_models, roi_models, ife_models)

diagnostic_rows <- purrr::imap_dfr(
  all_models,
  function(model, model_id) {
    if (is.null(model)) {
      return(tibble::tibble())
    }
    tibble::tibble(
      model_id = model_id,
      nobs = stats::nobs(model),
      singular = lme4::isSingular(model),
      aic = AIC(model),
      bic = BIC(model)
    )
  }
)

if (nrow(diagnostic_rows) > 0) {
  readr::write_csv(diagnostic_rows, file.path(output_dir, "mixed_model_diagnostics.csv"))
  diagnostic_rows %>% knitr::kable()
}
```

# Outputs

```{r output-list}
list.files(output_dir, recursive = TRUE) %>%
  tibble::tibble(output_file = .) %>%
  knitr::kable()
```

# Interpretation Checklist

Before writing claims:

1. Baseline bias must be reported before priming effects.
2. Priming should be reported as a shift away from no-prime baseline.
3. Filler baseline should be shown as a robustness comparison.
4. CORE/Jabberwocky differences should be interpreted only after token-count controls.
5. `model_size` is the primary model-comparison variable.
6. `model_family` should be descriptive or robustness-only.
7. ROI results should say whether effects happen at frame onset or later passive-cue regions.
8. IFE results should be framed as evidence for or against surprisal-linked adaptation, not assumed.
